---
title: "Recidivism risk assessment"
author: "Arvita, Marta, Saana, Polina, Tintti"
date: "June 27, 2018"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyr)
library(stargazer)
library(cowplot)

(WD <- getwd())
if (!is.null(WD)) setwd(WD)

recid <- read.csv('./DreselFarid2018/BROWARD_ORIGINAL.csv', head = TRUE, stringsAsFactors=FALSE)

chargeid <- read.csv('./DreselFarid2018/CHARGE_ID.csv', head = TRUE, stringsAsFactors=FALSE)

chargedesc <- read.csv('./DreselFarid2018/MTURK_CHARGE_DESC.csv', head = TRUE, stringsAsFactors=FALSE)

recid_subset <- read.csv('./DreselFarid2018/BROWARD_CLEAN_SUBSET.csv', head = TRUE, stringsAsFactors=FALSE)

mturk_norace <- read.csv('./DreselFarid2018/MTURK_NO_RACE.csv', head = TRUE, stringsAsFactors=FALSE)

mturk_race <- read.csv('./DreselFarid2018/MTURK_RACE.csv',head = TRUE, stringsAsFactors=FALSE)

mturk_norace_demo <- read.csv('./DreselFarid2018/MTURK_NO_RACE_DEMOGRAPHICS.csv', head = TRUE, stringsAsFactors=FALSE)

mturk_race_demo <- read.csv('./DreselFarid2018/MTURK_RACE_DEMOGRAPHICS.csv', head = TRUE, stringsAsFactors=FALSE)

#
mturk_norace <- mturk_norace[-1,]
names(mturk_norace)[1] <- "id"

mturk_race <- mturk_race[-1,]
names(mturk_race)[1] <- "id"

recid_subset$id <- as.character(recid_subset$id)
names(recid_subset)[10] <- "charge_degree"

### merge recid with charge IDs

merge_recid_chargeid <- full_join(recid, chargeid, by = "c_charge_desc")
write.csv(merge_recid_chargeid, 'C:/Users/mkolc/Google Drive/CONSIRT priv/Harmonia priv/Conferences, Workshop, Funding, Publications/SICSS Duke University June 2018/group project algorithmic fairness/DreselFarid2018/merge_recid_chargeid.csv')

merge_recid_chargeid_subset <- left_join(recid_subset, chargeid, by = "charge_id")
write.csv(merge_recid_chargeid_subset, 'C:/Users/mkolc/Google Drive/CONSIRT priv/Harmonia priv/Conferences, Workshop, Funding, Publications/SICSS Duke University June 2018/group project algorithmic fairness/DreselFarid2018/merge_recid_chargeid_subset.csv')

merge_recid_chargeid_subset_charge <- left_join(merge_recid_chargeid_subset, chargedesc[,c(2,3)], by = "mturk_charge_name")
write.csv(merge_recid_chargeid_subset_charge, 'C:/Users/mkolc/Google Drive/CONSIRT priv/Harmonia priv/Conferences, Workshop, Funding, Publications/SICSS Duke University June 2018/group project algorithmic fairness/DreselFarid2018/merge_recid_chargeid_subset_c.csv')

```
We used a database of over 7000 pretrial defendants arrested in 2013-2014 in Broward County, Florida. These individuals were matched with COMPAS scores. This dataset was analyzed in the ProPublica study, and then also used in the article by Dressel and Farid (2018). 

J. Dressel, H. Farid, The accuracy, fairness, and limits of predicting recidivism. Science Advances 4, eaao5580 (2018).

J. Angwin, J. Larson, S. Mattu, L. Kirchner, "Machine bias: There's software used across the country to predict future criminals. 
And it's biased against blacks," ProPublica, 23 May 2016; www.propublica.org/article/machine-bias-risk-assessmentsin-criminal-sentencing.

We started with a comparison of the racial composition of the dataset of offenders from Broward County, Florida, with the demographics from the whole county (Wikipedia, https://en.wikipedia.org/wiki/Broward_County,_Florida).
```{r}
Darthmouth.data <- round(prop.table(table(recid$race)),3)
Census.data <- c(0.177, 0.032, 0.425, 0.269, 0.004, 0.037)

rbind(Darthmouth.data, Census.data)
```

```{r}
mturk_norace_recid <- full_join(mturk_norace, recid_subset[,c(2,12)], by = "id")

mturk_norace_recid2 <- mturk_norace_recid %>% 
  mutate_if(is.numeric, funs(if_else(two_year_recid == 1, ., -.))) %>%
  mutate_if(is.numeric, funs(replace(., . == -1, 0))) %>%
  select(starts_with("X")) %>%
  mutate(share_positive_norace = rowSums(., na.rm = TRUE) / 20) %>%
  cbind(., recid_subset) %>%
  select(-starts_with("X")) %>%
  mutate(positive_decile_norace = ntile(share_positive_norace, 10))

mturk_race_recid <- full_join(mturk_race, recid_subset[,c(2,12)], by = "id")

mturk_race_recid2 <- mturk_race_recid %>% 
  mutate_if(is.numeric, funs(if_else(two_year_recid == 1, ., -.))) %>%
  mutate_if(is.numeric, funs(replace(., . == -1, 0))) %>%
  select(starts_with("X")) %>%
  mutate(share_positive_race = rowSums(., na.rm = TRUE) / 20) %>%
  cbind(., recid_subset) %>%
  select(-starts_with("X")) %>%
  mutate(positive_decile_race = ntile(share_positive_race, 10))

mturk_recid_all <- full_join(mturk_norace_recid2, mturk_race_recid2[,c(1,3,14)], by = "id")
```
Next we moved to analyzing data from the Dressel and Farid (2018) study. In this study the authors selected 1000 offenders from the Broward Country dataset, and divided them into 20 sets of 50. In an MTurk survey, these 20 sets of descriptions were shown to around 20 survey respondents each. Altogether the survey was filled out by over 400 respondents.

For each offender, 20 survey respondents evaluated whether the offender is likely to commit another crime in the next two years.
The respondents were told the offenders sex, age, the current charge and its degree, and the number of previous juvenile and non-juvenile charges.
For the same offenders, another 20 respondents evaluated whether the offender is likely to commit another crime in the next two years, with additional information about the offender's race.
We calculated the proportion of positive answers (i.e., proportion of "will reoffend" answers) for each offender, separately for the survey with and without race.
The distribution of the proportions of positive answers in both surveys (with and without race) is bimodal: for most offenders the proportion of positive answers is either less than 0.2 or over 0.8. This suggests that there is a group of offenders with medium risk that are difficult to evaluate by survey respondents.

```{r}
hist_norace <- ggplot(mturk_recid_all, aes(x = share_positive_norace)) + geom_histogram(breaks=seq(0, 1, by =0.1)) + 
  ggtitle("Survey without race") + xlab("")

hist_race <- ggplot(mturk_recid_all, aes(x = share_positive_race)) +  geom_histogram(breaks=seq(0, 1, by =0.1)) + 
  ggtitle("Survey with race")+ xlab("Proportion of positive answers")

cowplot::plot_grid(hist_norace, hist_race, ncol=1)
```
We then ranked the proportions of positive answers into deciles, separately for the survey with and without race. 
We calculated the Spearman's rank correlation coefficient for the risk deciles from COMPAS and the risk deciles from the survey with and without race. The correlation of survey decile ranks with the COMPAS deciles is 0.53 for the survey without race and 0.54 for the survey with race.

```{r}
round(cor(mturk_recid_all[,c("compas_decile_score", "positive_decile_race", "positive_decile_norace")], method = "spearman"),3)
```
```{r}
```
We also wanted to check the predictive power of the survey responses (from the survey with and without race) and the COMPAS decile for the risk of recidivism.
To do this, we estimated three logistic regression models predicting recidivism. The first model predicts recidivism with the risk decile from the survey without race. The second model predicts recidivism with the risk decile from the survey with race.
The second model predicts recidivism with the COMPAS decile. All three models are very similar in terms of magnitude of the effect and predictive power. This means that the survey responses are similarly effective in predicting recidivism as the COMPAS score.

```{r}
modela <- glm(two_year_recid ~ positive_decile_norace ,
             family=binomial(link='logit'), data=mturk_recid_all)

modelb <- glm(two_year_recid ~ positive_decile_race ,
             family=binomial(link='logit'), data=mturk_recid_all)

modelc <- glm(two_year_recid ~ compas_decile_score ,
              family=binomial(link='logit'), data=mturk_recid_all)

stargazer(modela, modelb, modelc, type = "text", title="Results", align=TRUE)

```
```{r}
```
We next ran OLS models predicting the decile score in surveys and from COMPAS with a set of variables about offenders: their sex, race, age, prior juvenile felony count, prior juvenile misdemeanor count, count of prior non-juvenile charges, and degree of current charge (felony or misdemeanor).
The predictive power of these offender characteristics is greater for the models predicting survey-based risk deciles than in the model predicting COMPAS deciles.

```{r}
mturk_recid_all$race <- as.factor(mturk_recid_all$race)

model1 <- lm(positive_decile_norace ~  race + sex + age + juv_fel_count + juv_misd_count + priors_count + charge_degree ,
             data=mturk_recid_all)

model2 <- lm(positive_decile_race ~ race + sex + age + juv_fel_count + juv_misd_count + priors_count + charge_degree,
             data=mturk_recid_all)

model3 <- lm(compas_decile_score ~ race + sex + age + juv_fel_count + juv_misd_count + priors_count + charge_degree,
             data=mturk_recid_all)

stargazer(model1, model2, model3, type = "text", title="Results", align=TRUE)
```
